%===============================================================================================================================%
%                                                            PREAMBLE                                                           %
%===============================================================================================================================%

\documentclass[honours,12pt]{unswthesis}
\linespread{1}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{latexsym,amsmath}
\usepackage{graphicx}
\usepackage{afterpage}
\usepackage{cite}
\usepackage{listings}
\usepackage{xcolor}
\usepackage[many]{tcolorbox}

% Macros
\newcommand{\N}{\mathbb{N}} % Natural numbers
\renewcommand{\P}{\mathbb{P}} % Prime numbers
\newcommand{\Z}{\mathbb{Z}} % Integers
\newcommand{\Q}{\mathbb{Q}} % Rational numbers
\newcommand{\A}{\mathbb{A}} % Algebraic numbers
\newcommand{\B}{\mathbb{B}} % Algebraic integers
\newcommand{\R}{\mathbb{R}} % Real numbers
\newcommand{\C}{\mathbb{C}} % Complex numbers
\newcommand{\T}{\mathbb{T}} % Circle group
\newcommand{\F}{\mathbb{F}} % Field

\newcommand{\disc}{\triangle}

\newcommand{\pmat}[1]{\begin{pmatrix}#1\end{pmatrix}}
\newcommand{\bmat}[1]{\begin{bmatrix}#1\end{bmatrix}}

\newcommand{\lnote}[1]{\llap{$#1$\hspace{0.75cm}}}
\newcommand{\rnote}[1]{\rlap{\hspace{0.75cm}$#1$}}

\newcommand{\red}[1]{\textcolor{red}{#1}}
\newcommand{\todo}[1]{\red{to do: #1}}
\newcommand{\tocheck}{(\red{check})}

\newcommand{\blankpage}{\null\thispagestyle{empty}\addtocounter{page}{-1}\newpage}


% Environments
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{conjecture}[theorem]{Conjecture}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{question}[theorem]{Question}
\newtheorem{notation}[theorem]{Notation}
\numberwithin{equation}{section}


% Code
\definecolor{codeblue}{rgb}{0, 0, 0}
\definecolor{backgroundcolor}{rgb}{0.8, 0.8, 0.8}

\lstdefinestyle{pystyle}{
    basicstyle=\ttfamily\footnotesize,
    breakatwhitespace=true,
    captionpos=b,
    keepspaces=true,
    numbers=left,
    numbersep=5pt,
    showspaces=false,
    showstringspaces=false,
    showtabs=false,
    tabsize=2
}

\newtcolorbox{pybox}{
    spartan,
    frame empty,
    boxsep = 0mm,
    left = 1mm,
    right = 1mm,
    top = -1mm,
    bottom = -1mm,
    colback = backgroundcolor
}

\lstset{style=pystyle}

\newcommand{\pyfile}[2][3]{
    \begin{pybox}
    \lstinputlisting[language=python, firstline=#1]{#2}
    \end{pybox}
}


%===============================================================================================================================%
%                                                            PREFACE                                                            %
%===============================================================================================================================%

% Title page
\title{Factoring via Binary Quadratic Forms}
\authornameonly{Harrison Graham}
\author{\Authornameonly\\{\bigskip}Supervisor: Professor David Harvey}
\copyrightfalse
\figurespagefalse
\tablespagefalse


% Plagiarism statement
\begin{document}
\beforepreface

\prefacesection{Plagiarism statement}
\bigskip \noindent I declare that this thesis is my
own work, except where acknowledged, and has not been submitted for
academic credit elsewhere. 

\bigskip \noindent I acknowledge that the assessor of this
thesis may, for the purpose of assessing it:
\begin{itemize}
\item Reproduce it and provide a copy to another member of the University; and/or,
\item Communicate a copy of it to a plagiarism checking service (which may then retain a copy of it on its database for the purpose of future plagiarism checking).
\end{itemize}

\bigskip \noindent I certify that I have read and understood the University Rules in
respect of Student Academic Misconduct, and am aware of any potential plagiarism penalties which may 
apply.\vspace{24pt}

\bigskip \noindent By signing 
this declaration I am
agreeing to the statements and conditions above.

\bigskip \noindent
Signed: \rule{7cm}{0.25pt} \hfill Date: \rule{4cm}{0.25pt} 


% Acknowledgements
% \pagebreak
% \prefacesection{Acknowledgements}
% \bigskip \noindent Body text


% Abstract
% \pagebreak
% \prefacesection{Abstract}
% \bigskip \noindent Body text

\afterpreface


%===============================================================================================================================%
%                                                           CHAPTER 1                                                           %
%===============================================================================================================================%

\pagebreak
\chapter{Introduction}\label{s-intro}

\red{to do}


%===============================================================================================================================%
%                                                           CHAPTER 2                                                           %
%===============================================================================================================================%

\pagebreak
\chapter{Binary Quadratic Forms}\label{s-forms}

A \textbf{binary quadratic form} (or simply a \textbf{form}) is a function of the form
$$f(x, y) = ax^2 + bxy + cy^2$$

It is called ``binary'' because it is a function of two variables, and ``quadratic'' because the degree of each term is 2.
For our purposes, we are usually working over the integers. 
Since $x$ and $y$ are variables, the behaviour of a given form is wholly dependent on the values of $a$, $b$ and $c$,
which motivates a shorthand notation of the same form; $(a, b, c)$.

We will impose that $a, c \neq 0$, or else $x$ or $y$ would be reduced to a linear factor.
Allowing this would serve no purpose beyond comlicating proofs with the added work of edge case considerations.

A form is called \textbf{primitive} if $\text{gcd}(a, b, c) = 1$. 
Primitive forms are of primary concern, as he behaviour of a nonprimitive form can be more simply understood by removing the common factor to work with simpler coefficients.


\section{Representation}\label{s-forms-rep}

A form $f(x, y)$ is said to \textbf{represent} a number $n$ if there exists an integer solution
$$f(x_0, y_0) = n \rnote{x_0, y_0 \in \Z}$$

For example, consider the form $(2, 1, 5)$, or $f(x, y) = 2x^2 + xy + 5y^2$. 
This form represents the number 2, since $f(1, 0) = 2$.
The form does not represent the number 1, though it is not immediately clear how to prove a result of this kind.

The questions which historically motivated the study of forms were as follows;
which numbers can be represented by a given form and, conversely, which forms can represent a given number?
These questions will not be our primary concern, but they will help to motivate the development of the theory that we need.


\section{Discriminant}\label{s-forms-disc}

The form $x^2 + y^2$ clearly exclusively represents non-negative integers. 
The same property holds for the previous example, $2x^2 + xy + 5y^2$, though this is harder to see.
On the other hand, the form $x^2 - y^2$ clearly represents both positive and negative integers.
A reasonable first step to understanding a form's representations is to classify whether they can be positive, negative, or both.

To do this, let us consider a form as a continuous real-valued function, and seek to find its zeroes
$$f(x, y) = ax^2 + bxy + cy^2 = 0$$

Notice that we can parameterise this equation by dividing by $y^2$ and setting $t = \tfrac{x}{y}$.
Then we can define
\begin{gather*}
    g(t) = at^2 + bt + c \\
    g(\tfrac{x}{y}) = \frac{f(x, y)}{y^2}
\end{gather*}

This may seem problematic for finding roots of the form $(x_0, 0)$, 
but substituting $y = 0$ into $f$ makes it clear that the only such zero is $(0, 0)$.
Except for this line, it is clear that $t(\tfrac{x}{y})$ agrees with $f(x, y)$ on zeroes and sign.
Hence, consider
\begin{gather*}
    at^2 + bt + c = 0 \\
    t = \frac{-b \pm \sqrt{b^2 - 4ac}}{2a}
\end{gather*}

If $b^2 -4ac$ is negative, then $g$ has no real zeroes, 
so $g$ (and hence $f$) must be either positive everywhere or negative everywhere (except the origin).
We call such forms \textbf{positive definite} and \textbf{negative definite}, respectively.
If $b^2 - 4ac$ is positive, then $g$ has two distinct real zeroes, so $g$ (and hence $f$) can be either positive or negative.
Such forms are called \textbf{indefinite}.
Similar reasoning shows that forms with $b^2 -4ac = 0$ are strictly non-negative or non-positive, 
but such forms are of little intereest.

This motivates defining the \textbf{discriminant} of a form $ax^2 + bxy + cy^2$ to be $\disc = b^2 - 4ac$.
We will come to see that this classification is far more powerful than simply being able to classify forms as definite/indefinite.

\todo{show that discriminants must be 0 or 1 mod 4}


\section{Equivalence}\label{s-forms-eq}
Given a form $f(x, y) = ax^2 + bxy + cy^2$, one can generate new forms which represent the same set of integers by changing the input space;
take some invertible function $\phi : \Z^2 \rightarrow \Z^2$, and let $g(x, y) = f \circ \phi(x, y)$.
Then for any $x_0, y_0 \in \Z$
$$ f(x_0, y_0) = n \implies g(\phi(x_0, y_0)) = n $$
$$ g(x_0, y_0) = m \implies f(\phi^{-1}(x_0, y_0)) = m $$

That is, all integers represented by $f$ are also represented by $g$, and vice versa. 
However, we must also ensure that our choice of $\phi$ gives $g$ to be a form.
Consider the choice $\phi(x, y) = (x+1, y)$.
This is an invertible function on $\Z^2$, but when it is composed with our usual choice of $f$ we have
\begin{align*}
    g(x, y) &= f \circ \phi(x, y) \\
    &= f(x+1, y) \\
    &= a(x+1)^2 + b(x+1)y + cy^2 \\
    &= ax^2 + 2ax + a + bxy + by + cy^2
\end{align*}

This is no longer a quadratic form, so this choice of $\phi$ should not be allowed. 
Instead, consider a function of the form $\phi(x, y) = (px+qy, rx+sy)$.
This gives
\begin{align*}
    g(x, y) &= f \circ \phi(x, y) \\
    &= f(px+qy, rx+sy) \\
    &= a(px+qy)^2 + b(px+qy)(rx+sy) + c(rx+sy)^2 \\
    &= \left[ap^2 + bpr + cr^2 \right] x^2
        + \left[ 2apq + b(ps + qr) + crs \right] xy
        + \left[ aq^2 + bqs + cs^2 \right] y^2
\end{align*}

This gives $g$ to be a binary quadratic form, as required. 
We now require that $\phi$ is invertible, so that this is a symmetric relationship.
Consider that $\phi$ can be understood as a matrix multiplication
$$\phi \left( \bmat{x \\ y} \right) = \bmat{p & q \\ r & s} \bmat{x \\ y}$$

So it is clear to see that the valid choices of $\phi$ are the invertible elements of $M_2(\Z)$. 
These are the elements of $GL_2(\Z)$, also known as the \textbf{unimodular} matrices (of size 2).
The determinant of such a matrix will always be $\pm 1$ 
(since the determinant is a result of sums and products of the integer coefficients, 
and the product of determinants of a matrix and its inverse is 1).

We define forms to be \textbf{improperly equivalent} if there is a $GL_2(\Z)$ transformation between them.
It is easy to see that this is an equivalence realtion.
If we restrict the discriminant of the transformation to be strictly positive 1, we split the equivalence classes further-
we call such forms \textbf{properly equivalent}, which is also an equivalence relation.


\section{Reduction}\label{s-forms-red}
Now that we have an equivalence relation, we would like to be able to find a suitable representative element of a given class.
This will not only prove useful for simplifying computations by choosing a representative with small values of $a$, $b$, and $c$,
but will also give us a method for checking whether two forms belong to the same class by evaluating and comparing their canonical representatives.

A basis for $SL_2(\Z)$ is given by
$$S = \pmat{0 & -1 \\ 1 & 0}, \hspace{5mm} T = \pmat{1 & 1 \\ 0 & 1}$$

Note that
$$S^4 = I, \hspace{5mm} T^k = \pmat{1 & k \\ 0 & 1}$$

So we see that any properly equivalent forms must be related by some series of applications of $S$ and $T^k$.
Let us observe the results of these transformations on a generic form $f(x, y) = (a, b, c)$
\begin{align*}
    f \circ S (x, y) &= f \circ \pmat{0 & -1 \\ 1 & 0} \pmat{x \\ y} \\
    &= f(-y, x) \\
    &= a(-y)^2 + b(-y)x + cx^2 \\
    &= ay^2 - bxy + cx^2
\end{align*}
\begin{align*}
    f \circ T^k (x, y) &= f \circ \pmat{1 & k \\ 0 & 1} \pmat{x \\ y} \\
    &= f(x + ky, y) \\
    &= a(x + ky)^2 + b(x + ky)y + cy^2 \\
    &= a(x^2 + 2kxy + k^2y^2) + b(xy + ky^2) + cy^2 \\
    &= ax^2 + (2ak + b)xy + (ak^2 + bk + c)y^2
\end{align*}

So we have
\begin{gather*}
    S : (a, b, c) \mapsto (c, -b, a) \\
    T^k : (a, b, c) \mapsto (a, 2ak + b, ak^2 + bk + c)
\end{gather*}

Let us use these for constructing a redution algorithm, which turns any form into a canonical class representative.
For now, we will only consider positive definite forms, as the result is simpler.
Note that in this case, $a$ and $c$ must both be positive, as their signs must be equal for $\disc = b^2 - 4ac < 0$,
and they must be positive to represent positive integers.

A natural way to use $S$ and $T$ to reduce a form is to attempt to minimise $|b|$.
Whenever $|b| > a$, we can use $T$ to reduce $b$ to be in the range $[-a, a]$.
Then, if $c < a$, we can use $S$ to swap $a$ and $c$, then return to step 1 to reduce $b$ further.

Eventually, we reach a form with $|b| \leq a \leq c$.
This algorithm can easily be shown to terminate, as $|b|$ decreases with each repetition, so there can be at most $|b|$ repetitions.

\todo{fix this part}
In the case that $a = |b|$ or $a = c$, the above criteria is satisfied by both $(a, b, c)$ and $(a, -b, c)$, 
so we will choose $(a, |b|, c)$ to be the canonical representative.

\pyfile[1]{algorithms/definite-reduction.py}

\todo{prove the reduced representatives are canonical}

\todo{highlight that equivalent forms have the same discriminant}


%===============================================================================================================================%
%                                                           CHAPTER 3                                                           %
%===============================================================================================================================%

\chapter{Composition}\label{s-comp}
Now that we have the means to properly understand equivalence classes of forms, we seek to equip them with a binary operation to form a group.
We acheive this by showing a bijection between classes of forms and classes of ideals, 
under which the ideal multiplication structure gives a multiplicative structure to the corresponding classes of forms.


\section{Ideal Correspondence}\label{s-comp-ideals}
As was established in \ref{s-forms-disc}, a form $(a, b, c$) when considered as a continuous function has zeroes at
$$t = \frac{x}{y} = \frac{b^2 \pm \sqrt\disc}{2a}$$

From this, we can define a ring given by
$$\Z\left[\frac{b^2 \pm \sqrt\disc}{2a}\right] = \Z + \frac{b^2 - \sqrt\disc}{2a}\Z$$

But this isn't quite the formulation that we need - it's not immediately clear how to understand these rings as ideals of some common larger ring.
However, if we multiply by $a$, we get a much more useful result:
$$a\Z + \frac{b^2 - \sqrt\disc}{2}\Z$$

A ring of this form is an ideal of the ring
$$\Z\left[\frac{b - \sqrt\disc}{2}\right] = 
\begin{cases}
    b \equiv 0 \hspace{2mm} \text{(mod 2)} & \Z + \frac{0 - \sqrt\disc}{2} \\
    b \equiv 1 \hspace{2mm} \text{(mod 2)} & \Z + \frac{1 - \sqrt\disc}{2}
\end{cases}$$

Noting that $\disc = b^2 - 4ac$ requires $\disc \equiv b^2 \hspace{2mm} \text{(mod 4)}$, 
then we also have $\disc \equiv b \hspace{2mm} \text{(mod 2)}$.
So the ideals we defined are contained within
$$\Z + \frac{\disc - \sqrt\disc}{2}\Z$$

This depends only on the discriminant, so forms of the same discriminant correspond to ideals of the same ring, which we will denote $\mathcal{O}_\disc$.
We have defined a map from forms of discriminant $\disc$ to ideals of $\mathcal{O}_\disc$
$$\varphi : (a, b, c) \mapsto a\Z + \frac{b - \sqrt\disc}{2}\Z$$

For this to give any meaningful structure, we must show that equivalence of forms corresponds to some notion of equivalence of ideals.
Consider how $\varphi$ acts on forms which vary by an action of $T$
\begin{align*}
    \varphi \circ f \circ T(x, y) &= \varphi(ax^2 + (2a + b)xy + (a + b + c)y^2) \\
    &= a\Z + \frac{2a + b - \sqrt\disc}{2} \\
    &= a\Z + \frac{b - \sqrt\disc}{2} \\
    &= \varphi \circ f(x, y)
\end{align*}

So we see that $\varphi$ is invariant under actions of $T$.
Now consider actions of $S$
\begin{align*}
    \varphi \circ f \circ S(x, y) &= \varphi(cx^2 -bxy + ay^2) \\
    &= c\Z + \frac{-b - \sqrt\disc}{2} \\
\end{align*}


\todo{finish}


%===============================================================================================================================%
%                                                           CHAPTER 4                                                           %
%===============================================================================================================================%

\chapter{Factoring}\label{s-factor}
A key practical application of number theory is the factoring algorithms that its theory allows.
Factoring is a deceptively difficult task to perform efficiently.
As a baseline for comparison, let us consider the most primitive factoring algorithm - trial division.

\pyfile[1]{algorithms/trial-division.py}

Note that all code samples will be provided in python, using numpy for mathematical functions - 
the import will be omitted going forward.

We will use big-$O$ notation to analyse the complexity of factoring algorithms.
It might seem natural to class this algorithm as $O(\sqrt{n})$, but the standard convention for measuring the complexity
of factoring algorithms is not that $n$ measures the size of the number being factored -
rather, it measures the size of the number being factored, in binary digits.
Therefore, the time complexity of the trial division algorithm is $O(2^{n/2})$
\red{correct: "exponential" notation}


\section{Ambiguous forms}\label{s-factor-amb}
The key insight which bridges the theory of binary quadratic forms into the realm of factoring comes from considering
forms of order 2 - that is, forms which are their own inverses. 
Such forms are called \textbf{ambiguous forms}, since they are properly equivalent to the same forms that they are improperly
equivalent to, so it is ``ambiguous'' which proper equivalence class they belong to.

For a reduced form to be properly equivalent to its inverse, we have
$$(a, b, c) = (a, -b, c)$$

When is this possible?
The clearest case where the equility holds is when $b = 0$, but there are other possibilities.
If $b \neq 0$, the only way for the equivalence to hold is if $(a, -b, c)$ is not fully reduced.
Since $(a, b, c)$ is reduced, we know that $|b| \leq a \leq c$, 
so the only way for $(a, -b, c)$ not to be reduced is if $-b < 0$ and $a = |-b|$ or $a = c$.
Since $(a, b, c)$ is reduced, either of the latter conditions would imply $b > 0$, so we can now write the three cases
\begin{gather*}
    \textbf{(1)} \hspace{2mm} b = 0 \implies (a, 0, c) \\
    \textbf{(2)} \hspace{2mm} a = b \implies (a, a, c) \\
    \textbf{(3)} \hspace{2mm} a = c \implies (a, b, a)
\end{gather*}

An ambiguous form must necessarily meet one of these conditions, and it is easy to show that each consition is sufficient:
\begin{align*}
    \textbf{(1)} \hspace{2mm}
    (a, 0, c) \cdot (a, 0, c) &= (a, 0, c) \cdot (a, -0, c) \\
    &= id_\disc
\end{align*}
\begin{align*}
    \textbf{(2)} \hspace{2mm}
    (a, a, c) \cdot (a, a, c) &= (a, a, c) \cdot (a, -a, a - a + c) \rnote{\text{using $T^{-1}$}} \\
    &= (a, a, c) \cdot (a, -a, c) \\
    &= id_\disc
\end{align*}
\begin{align*}
    \textbf{(3)} \hspace{2mm}
    (a, b, a) \cdot (a, b, a) &= (a, b, a) \cdot (a, -b, a) \rnote{\text{using $S$}} \\
    &= id_\disc
\end{align*}

So these cases are necessary and sufficient to classify a form as ambiguous.
Notice that each case reduces the three variables $a, b, c$ to only two - 
this is of particular note when one considers how each case affects the formula $\disc = b^2 - 4ac$
\begin{gather*}
    \textbf{(1)} \hspace{2mm} \disc = 0 - 4ac = -4ac \\
    \textbf{(2)} \hspace{2mm} \disc = a^2 - 4ac = a(a-4c) \\
    \textbf{(3)} \hspace{2mm} \disc = b^2 - 4a^2 = (b + 2a)(b - 2a)
\end{gather*}

Notice that every case gives a factorisation of the discriminant.
Therefore, if we want to find a factorisiation of some number $D$, we can do this by finding amabiguous forms of discriminant $\disc = D$
(or $\disc = 4D$ if $D$ is congruent to 2 or 3 modulo 4).

\todo{add references}
\cite[Ch.\,3]{buchmann}
\cite[Thm.\,III.3.5]{buchmann}

\todo{add roadmap}


%===============================================================================================================================%
%                                                           REFERENCES                                                          %
%===============================================================================================================================%

\pagebreak
\addcontentsline{toc}{chapter}{References}
\bibliographystyle{amsalpha}
\bibliography{references}
\end{document}